Type-Safe Modular Parsing

### 2016-12-26 paper/sections/CaseStudy.tex
Some more comments on Case studies and Implementation

@@

To demonstrate the utility of our modular parsing library, we implemented parsers of the first 18 calculi from book \textit{Types and Programming Languages}. In the book it introduces several calculi from simple to complex, by gradually adding new features to syntax. It is suitable for our case study for mainly two reason. First, the calculi are arguably practical as we want to show our library could be used in real world development. One of the most complex calculi in our case study is System F (with several extensions such as records and variants), which could be the foundation of practical functional programming languages. Second, the evolution of calculi in the book reveals the advantages of modular representation of abstract syntax and modular parsing, which is the key functionality of our library. By decomposing those calculi into smaller components and reusing the common ones, we can obtain considerably code reuse in parsing with little performance penalty, as shown later.

Using the modular parsing techniques discussed before, we extract reusable \textit{components} from syntax of all the calculi. Each component, which may contain several syntactical structures, is a reusable fragment representing a certain feature in the syntax. For example, \lstinline{VarApp} component represents the variable case and application case of expressions. In our case study, each component is represented by a Scala object which has three parts. \lstinline{Alg} is an object algebra interface for the abstract syntax. \lstinline{Print} is an object algebra implements that interface \lstinline{Alg}, for the pretty printing operation. It is not necessary for parsing, but we include it here for demonstration. \lstinline{Parser} is the modular parser for this piece of syntax.

@@

Each calculus could be composed directly from components and other calculi, if it has those common structures in the syntax. For example, the calculus \lstinline{Untyped}, representing the famous untyped lambda calculus, can be constructed from components \lstinline{VarApp} and untyped lambda abstraction \lstinline{UntypedAbs}.

@@

We compared our implementation of the parsers with an implementation we found online, written by author Ilya Klyuchnikov. Ilya's implementation is very suitable for comparison, because it is also written in Scala using the same parser combinator library. Furthermore, it includes parsers of all the 18 calculi we have, but written in a non-modular way. Thus it is not able to reuse existing code of parsing common syntactical features.

@@

The comparison is made from two aspects. First, we want to show that using our modular parsing library, the code can be reused in a significant way. For this propose, we measured source lines of code (SLOC) of the two implementations. All blank lines and comment lines are excluded, and we formatted the code of both implementations so that the length of each line does not exceed 120 characters. Second, users may concern the performance penalty caused by modularity. Thus we compared the execution time, as a major measurement of performance, between the two implementations. We prepared test files containing valid expressions in each calculus, then we used the corresponding parser to parse the file and the pretty printer to print the result. The execution time excluding reading input file was calculated.

@@

Table \ref{tab:comparison} shows data of the comparison. The overall result of SLOC comparison is that 66.3\% of code is reduced using our library. That is because the code of some common syntax structures such as variable, lambda abstraction, etc., is reused several times in the whole project. We can see that the first two calculi \lstinline{Arith} and \lstinline{Untyped} we do worse than Ilya's implementation, because in such two cases we do not reuse any existing components. But in each of the following 16 calculi, we can compose our syntax definition, parser and pretty printer from extending reusable components, resulting considerably code reduction.

For the execution time comparison, we built a generator to randomly generate valid expressions according to the syntax of each calculus. Each test file corresponds with one calculus and it consists of 500 expressions randomly generated. The size of test files varies from 20KB to 100KB. After testing both implementation, we found that our version is about \huang{60\%} slower. Part of the reason is that our library requires more intermediate function calls, and the object algebra and self-reference needed to be passed as arguments every time. However, another important reason is that, because of the modular setting, we use a different alternative combinator which will try both branches and match the longest one. \huang{ref to somewhere before?} In contrast Ilya uses the ordinary alternative combinator which tries the second branch only if the first one fails. Although this difference slows our version, we want to mention that because of the memorization of Packrat parsers, we are just constant times slower, the big-O complexity is still the same.

>>>>

To demonstrate the utility of our modular parsing library, we
implemented parsers of the first 18 calculi from book \textit{Types
  and Programming Languages} (TAPL). TAPL introduces several
calculi from simple to complex, by gradually adding new features to
syntax. \bruno{mention that you compare against an existing
  non-modular implementation.}

@@

The calculi in the TAPL book are suitable for our case study for mainly two reasons.
Firstly, the calculi are arguably capture many of the language features 
required in realistic programming languages. Secondly, the evolution of
calculi in the book reveals the advantages of modular representation
of abstract syntax and modular parsing, which is the key functionality
of our library. By decomposing those calculi into smaller components
and reusing the common ones, we can obtain considerably code reuse in
parsing with little performance penalty, as shown later.

\paragraph{Extracting Language Components}
Using the modular parsing techniques discussed before, we extract
reusable \textit{components} from syntax of all the calculi. Each
component, which may contain several syntactical structures, is a
reusable fragment representing a certain feature in the syntax. For
example, \lstinline{VarApp} component represents the variable case and
application case of expressions. In our case study, each component is
represented by a Scala object which has three parts. \lstinline{Alg}
is an object algebra interface for the abstract syntax.
\lstinline{Print} is an object algebra implements that interface
\lstinline{Alg}, for the pretty printing operation. It is not
necessary for parsing, but we include it here for demonstration.
\lstinline{Parser} is the modular parser for this piece of syntax.

@@

\paragraph{Composing Language Components}
Each calculus could be composed directly from components and other
calculi, if it has those common structures in the syntax. For example,
the calculus \lstinline{Untyped}, representing the famous untyped
lambda calculus, can be constructed from components \lstinline{VarApp}
and untyped lambda abstraction \lstinline{UntypedAbs}.

@@

We compared our implementation of the parsers with an implementation
by  Ilya Klyuchnikov, available online. Ilya's
implementation is suitable for comparison, because it is also
written in Scala using the same parser combinator library.
Furthermore, it includes parsers of all the 18 calculi we have, but
written in a non-modular way. Thus it is not able to reuse existing
code of parsing common syntactical features.

@@

The comparison is made from two aspects. First, we want to show that
using our modular parsing library, the code can be reused in a
significant way. For this propose, we measured source lines of code
(SLOC) of the two implementations. All blank lines and comment lines
are excluded, and we formatted the code of both implementations so
that the length of each line does not exceed 120 characters. Second,
we are interested to assess the performance penalty caused by modularity. Thus
we compared the execution time, as a major measurement of performance,
between the two implementations. We prepared test files containing
valid expressions in each calculus,\bruno{say something about what
  those expressions are, size, and how they were generated.} then we used the corresponding
parser to parse the file and the pretty printer to print the result.
The execution time excluding reading input file was calculated.

@@

Table \ref{tab:comparison} shows data of the comparison. The overall
result of SLOC comparison is that 66.3\% of code is reduced using our
library. That is because the code of some common syntax structures
such as variables, lambda abstractions, etc., are reused several times in
the whole project. We can see that in the first two calculi
\lstinline{Arith} and \lstinline{Untyped} we do worse than Ilya's
implementation, because in such two cases we do not reuse any existing
components. But in each of the following 16 calculi, we can compose
our syntax definition, parser and pretty printer from extending
reusable components, resulting considerably code reduction.

\paragraph{Performance}
For the execution time comparison, we built a generator to randomly
generate valid expressions according to the syntax of each calculus.
Each test file corresponds with one calculus and it consists of 500
expressions randomly generated. The size of test files varies from
20KB to 100KB. After testing both implementation, we found that our
version is about \huang{60\%} slower. \bruno{A table showing a summary
of results is missing here. We discussed this before.}
 Part of the reason is that our
library requires more intermediate function calls, and the object
algebra and self-references need to be passed as arguments every
time. However, another important reason is that, because of the
modular setting, we use a different alternative combinator which will
try both branches and match the longest one. \huang{ref to somewhere
  before?} In contrast Ilya uses the ordinary alternative combinator
which tries the second branch only if the first one fails. Although
this difference slows our version, we want to mention that because of
the memoization of Packrat parsers, we are just constant times
slower, the big-O complexity is still the same.
